Increased productivity using Atomify - a real-time LAMMPS visualizer

### 2018-05-30 ms.tex
Specification to availability
bleh

\title{Atomify - a live LAMMPS visualizer}
\author{Anders Hafreager$^1$}
\author{Svenn-Arne Dragly$^{1}$} 
\author{Anders Malthe-S\o renssen$^1$}
\affiliation{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway }
\date{\today}

@@

\begin{abstract}
%
The typical LAMMPS workflow includes working with a text editor to modify
scripts, the terminal to run simulations, and programs like VMD or Ovito to
visualize trajectories.
If physical quantities are computed, the data is often plotted with MATLAB or
Python, where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who
are new to LAMMPS.
We here introduce Atomify;
a high performance live visualizer for LAMMPS simulations,
with stunning graphics able to simulate and render more than 250,000 atoms with
excellent frame rate on modern hardware.
Atomify supports OpenMP and GPU acceleration, live plotting of LAMMPS variables
and computes, and an easy-to-use code editor in one single program.
Direct access to the powerful machinery already built into LAMMPS allows easy
access to advanced physical quantities.
Atomify is open-source software (GPL) written in C++ using the Qt framework.
%
\end{abstract}

@@

\section{Introduction}
%
Molecular dynamics (MD) is a common technique for simulating the motion of atoms, molecules and larger particles in a broad range of fields such as physics, chemistry and biology \cite{flerefelt}.
% TODO does MD also include larger particles?
% TODO which fields?
Over the years, increasingly sophisticated methods have been developed.
% TODO what methods?
With the large number of force fields, advanced techniques and GPU acceleration,
implementing a good molecular dynamics simulation is a challenging task.
There are many software packages for molecular modelling
available\footnote{\url{https://en.wikipedia.org/wiki/Comparison_of_software_for_molecular_mechanics_modeling} compares different softwares for molecular modeling.}.
Some packages, such as LAMMPS\cite{Plimpton1995Fast}, GROMACS\cite{Pronk2013}
and NAMD\cite{Phillips2005Scalable} have been developed for more than two decades 
and contain more than one million lines of code each.

The work in this paper is focused on LAMMPS.
Although the software is well-documented and enables the user to perform
advanced simulations, it can be demanding to learn how to use it well.
A LAMMPS simulation is performed by a series of commands which are executed one by
one.
Physical quantities such as energy, temperature and stress can be measured by
adding \textit{computes}.
The time evolution of these quantities can in turn be saved to file for plotting
and further analysis.
Particle trajectories can be saved to a file and visualized by tools such as
VMD\cite{Humphrey1996Vmd} or OVITO\cite{Stukowski2009Visualization}.
The lack of an intuitive GUI and visualization\footnote{GUI is listed as a non-feature on the web page:
\url{http://lammps.sandia.gov/non_features.html}.} makes the workflow somewhat
complicated, see figure \ref{fig:flowchart}.

Atomify solves combines script editing, simulation,
visualization, and analysis in one single software application.
It uses LAMMPS as a library and synchronizes after each timestep,
copying the relevant information from LAMMPS to the graphics processing unit
(GPU) for visualization and data plotting.
The visualization is optimized for high performance by raytracing billboards.
This allows visualization of millions of atoms with good framerate on desktop
computers. Atomify is designed with simplicity as main goal.

@@

Certain Atomify-commands can be added to the comments of a regular LAMMPS script
to modify properties of the GUI.
All these start with \keys{\texttt{\#}} so that LAMMPS ignores them as they are
interpreted as comments.
For instance, the color and size of the atoms can be set based on the type by
the command \keys{\texttt{\#/atom}}.
To set the color to white and radius to 2.0 for particle type 1,
the user can apply the command \keys{\texttt{\#/atom 1 2.0 \#ffffff}}.
Certain built-in atom types have default values and can be accessed with
\keys{\texttt{\#/atom 1 oxygen}} for simplicity.
The initial camera position can also be set similarily.
For a full list of the available commands, see the Atomify
documentation\cite{atomifydocumentation}.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{figures/temperature.png}
	\caption{Atomify can visualize computes defined in LAMMPS. 
    Here the temperature is plotted as a function of time.}
	\label{fig:temperature}
\end{figure}

@@

Each sphere is visualized as two triangles making up a square large enough to
cover all pixels the sphere needs.
To reduce the CPU workload, all 4 corner vertices of the square are placed at
the exact position of the sphere.
The radius and a vertex id is specified, so that the vertex shader can move the
4 vertices so that the plane is orthogonal to the view vector and is larger than
the cross sectional area of the sphere of interest.
This is illustrated in figure \ref{fig:billboards}.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{spherebillboards.png}
	\caption{Illustration of billboards. The circles are shown in the figure to show our target rendering. a) all 4 vertices are at the center of the sphere. b) the vertex shader moves each vertex so that the plane covers a larger cross sectional area than the sphere. c) the orientation of the plane is such that the normal vector of the plane is parallel to the view vector. }
	\label{fig:billboards}
\end{figure}

@@

Atomify continues in heavy development and we have several features planned for
future versions.
A web version of Atomify using WebGL and WebAssembly will make it easy to test
scripts in the browser.
This is useful in educational settings where teachers wish to introduce students
to LAMMPS without installing additional software.
It will also serve as a nice demonstration of the full desktop application of
Atomify.

@@

% TODO Reference to KOKKOS
We also aim to include more LAMMPS backends, such as KOKKOS, and running
simulations on computing clusters while visualizing them on the users own
computer.

@@

Atomify is availabe on Linux, MacOS and Windows and can be downloaded from
\href{https://ovilab.net/atomify}{http://ovilab.net/atomify}.
A lightweight version of the desktop application is also available for mobile
devices running Android or iOS.

>>>>

\title{Increased productivity using Atomify - a real-time LAMMPS visualizer}
\author{Anders Hafreager$^1$}
\author{Svenn-Arne Dragly$^{1}$} 
\author{Anders Malthe-S\o renssen$^1$}
\affiliation{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway }
\date{\today}

@@

\begin{abstract}
%
The typical workflow when running atomistic simulations includes working with several programs.
A text editor is needed to create and modify the scripts, the terminal to run the simulation, and programs like VMD or Ovito to visualize the system over time.
If physical quantities are computed, the data is often plotted with MATLAB or Python, where additional scripts must be used.
This is a tedious process, especially for teaching purposes and for people who are new in the field.
We here introduce Atomify; a high performance real-time visualizer for atomistic simulations that can simulate and render more than one million atoms with excellent frame rate on modern hardware.
Atomify supports OpenMP acceleration, GPU acceleration, real-time plotting of physical quantities, and an easy-to-use code editor in one single application.
It currently uses LAMMPS as physics engine, but it can be extended to support other codes like Gromacs, NAMD or OpenMM.
Atomify is open-source software (GPL) written in C++ using the Qt framework.
%
\end{abstract}

@@

\section{Introduction}
%
Molecular dynamics (MD) is a common technique for simulating the motion of atoms,
molecules and larger particles in a broad range of fields such as physics, chemistry and biology \citep{van2001reaxff, allen2017computer, }.
% TODO does MD also include larger particles?
% TODO which fields?
Over the years, increasingly sophisticated methods have been developed.
% TODO what methods?
With the large number of force fields, advanced techniques and GPU acceleration,
implementing a good molecular dynamics simulation is a challenging task.
There are many software packages for molecular modelling
available\footnote{\url{https://en.wikipedia.org/wiki/Comparison_of_software_for_molecular_mechanics_modeling} compares different softwares for molecular modeling.}.
Some packages, such as LAMMPS\citep{Plimpton1995Fast}, Gromacs\citep{berendsen1995gromacs}
and NAMD\citep{Phillips2005Scalable} have been developed for more than two decades 
and contain more than one million lines of code each.

The work in this paper is focused on LAMMPS.
Although the software is well-documented and enables the user to perform
advanced simulations, it can be demanding to learn how to use it well.
A LAMMPS simulation is performed by a series of commands which are executed one by
one.
Physical quantities such as energy, temperature and stress can be measured by
adding \textit{computes}.
The time evolution of these quantities can in turn be saved to file for plotting
and further analysis.
Particle trajectories can be saved to a file and visualized by tools such as
VMD\citep{Humphrey1996Vmd} or OVITO\citep{Stukowski2009Visualization}.
The lack of an intuitive GUI and visualization\footnote{GUI is listed as a non-feature on the web page:
\url{http://lammps.sandia.gov/non_features.html}.} makes the workflow somewhat
complicated, see figure \ref{fig:flowchart}.

Atomify combines script editing, simulation,
visualization, and analysis in one single application.
It uses LAMMPS as a library and synchronizes after each timestep,
copying the relevant information from LAMMPS to the graphics processing unit
(GPU) for visualization and data plotting.
The visualization is optimized for high performance by raytracing billboards.
This allows visualization of millions of atoms with good framerate on desktop
computers. Atomify is designed with simplicity as main goal.

@@

Certain Atomify-commands can be added to the comments of a regular LAMMPS script
to modify properties of the GUI.
All these start with \keys{\texttt{\#}} so that LAMMPS ignores them as they are
interpreted as comments.
For instance, the color and size of the atoms can be set based on the type by
the command \keys{\texttt{\#/atom}}.
To set the color to white and radius to 2.0 for particle type 1,
the user can apply the command \keys{\texttt{\#/atom 1 2.0 \#ffffff}}.
Certain built-in atom types have default values and can be accessed with
\keys{\texttt{\#/atom 1 oxygen}} for simplicity.
The initial camera position can also be set similarily.
For a full list of the available commands, see the Atomify
documentation\citep{atomifydocumentation}.

@@

% \begin{figure}
% 	\centering
% 	\includegraphics[width=0.45\textwidth]{figures/temperature.png}
% 	\caption{Atomify can visualize computes defined in LAMMPS. 
%     Here the temperature is plotted as a function of time.}
% 	\label{fig:temperature}
% \end{figure}

@@

Each sphere is visualized as two triangles making up a square large enough to
cover all pixels the sphere needs.
To reduce the CPU workload, all 4 corner vertices of the square are placed at
the exact position of the sphere.
The radius and a vertex id is specified, so that the vertex shader can move the
4 vertices so that the plane is orthogonal to the view vector and is larger than
the cross sectional area of the sphere of interest.

@@

Atomify continues in heavy development and we have several features planned for
future versions.
We already have a simple version of Atomify in the web browser using WebGL and WebAssembly.
This makes it easy to test scripts in the browser with all simple functionalities available.
This is especially useful in educational settings where teachers wish to introduce students
to LAMMPS without installing additional software.
It will also serve as a nice demonstration of the full desktop application of
Atomify.

@@

% TODO Reference to KOKKOS
We also aim to include more LAMMPS backends, such as KOKKOS, and running
simulations on computing clusters while visualizing them on the users own
computers.

@@

Atomify is availabe on Linux, MacOS and Windows and can be downloaded from
\href{https://ovilab.net/atomify}{http://ovilab.net/atomify}.
A lightweight version of Atomify is also available for mobile devices running
Android or iOS.
